\ProjectName{}: Data Imputation as a Query Optimization

### 2016-12-14 paper/algorithm.tex
paper: Add some last edits

\subsection{Imputation operators}
We introduce two new relational operators to perform imputation: impute ($\mu$) and drop ($\delta$). Each operator takes arguments $(C, R)$ where $C$ is a set of attributes and $R$ is a relation. Impute uses a machine learning method to replace all null values with non-null values for attributes in $C$ in the relation $R$. Drop simply removes from $R$ all tuples which have a null value for some attribute in $C$. Both operators guarantee that the resulting relation will contain no null values for attributes in $C$.

@@

For each sub-query, using the histogram information, we keep track of the estimated number of
null values in a column  ($\text{MissingCount}(c, Q)$). We use this estimate to compute
$\text{Loss}(Q)$ as follows.

@@

@@

We design our system such that any imputation strategy can be plugged in with minimal effort
and without changes to the optimizer, so the strategy can in principle be targeted to a
specific domain. The current version of the system uses a general-purpose imputation
strategy based on chained-equation classification and regression trees (CE-CART). Chained
equation imputation methods \cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression} \cite{gelman2006data}) impute multiple missing attributes by
iteratively estimating predictive models of one missing attribute conditional on all
complete attributes and the other missing attributes. The individual predictive models can
be customized by the analyst to the problem at hand. Decision tree algorithms, like CART,
are found to be effective \cite{akande2015empirical} in empirical studies in general-purpose
routines and are widely used in epidemiological domains \cite{burgette2010multiple}.

These imputation algorithms are rarely used in isolation --- rather, the ultimate goal is to
perform statistical analysis of the complete data. To this end, the technique of
\textit{multiple imputation} can be used, in which multiple distinct copies of the complete
data are generated and estimators are computed by averaging over the multiple datasets. In
ImputeDB, we cannot assume the user desires multiple copies of the query result, though we
note that in further work, multiple imputation could be used within ImputeDB to generate
lower-variance estimates of some aggregates, like averages.

@@

It is worth noting that given the nature of the imputation strategy used here, imputation
operators become blocking in our system. An extension of ImputeDB could consider regression
algorithms that learn in an online nature; this would reduce the cost of the imputation at
the possible expense of quality.

@@

\subsection{Complexity}
Our optimization algorithm builds off the approach taken by System R\cite{blasgen1981system}, therefore our algorithm still operates in exponential time. Indeed, 
note that if we remove our restriction on types of imputation($\delta_{min}, \mu_{min}, \mu_{max}$), and allow any arbitrary subset of attributes to be imputed,
then every single operator in the query plan has a number of imputations exponential in the number of dirty columns. Our restriction, instead, increases the number
of plans (in the worst case) at each operator by a factor of 3. Of course, this implies that in the worst case (where all dirty sets tracked are distinct throughout the query plan),
we explore a number of plans that further scales the number of plans in the original algorithm by an exponential factor.

>>>>

\subsection{Imputation operators}
We introduce two new relational operators to perform imputation: \textit{Impute} ($\mu$) and
\textit{Drop} ($\delta$). Each operator takes arguments $(C, R)$ where $C$ is a set of
attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
replace all null values with non-null values for attributes in $C$ in the relation $R$.
\textit{Drop} simply removes from $R$ all tuples which have a null value for some attribute in $C$.
Both operators guarantee that the resulting relation will contain no null values for
attributes in $C$.

@@

For each sub-query, using the histogram information, we keep track of the estimated number of
null values in a column  ($\text{MissingCount}(c, Q)$), given the histograms associated with
the sub-query. We use this estimate to compute
$\text{Loss}(Q)$ as follows.

@@

\todo{is it necessary to add a fudge factor to the loss calculation, for generality, and set
it at 1.0 for our paper?}

@@

We design our system such that any imputation strategy can be plugged in with minimal effort
and without changes to the optimizer, so the strategy can in principle be targeted to a
specific domain. The current version of the system uses a general-purpose imputation
strategy based on chained-equation classification and regression trees (CE-CART). Chained
equation imputation methods \cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression} \cite{gelman2006data}) impute multiple missing attributes by
iteratively estimating predictive models of one missing attribute conditional on all
complete attributes and the other missing attributes. The individual predictive models can
be customized by the analyst to the problem at hand. Decision tree algorithms, like CART,
are found to be effective \cite{akande2015empirical} in empirical studies in general-purpose
routines and are widely used in epidemiological domains \cite{burgette2010multiple}. Though
we provide quantitative results for the CE-CART implementation, ImputeDB is designed to be
agnostic to the choice of imputation algorithm, and additional algorithms --- with
associated time and loss functions --- could be insert without any change in design.

These imputation algorithms are rarely used in isolation --- rather, the ultimate goal is to
perform statistical analysis of the complete data. To this end, the technique of
\textit{multiple imputation} can be used, in which multiple distinct copies of the complete
data are generated and estimators are computed by averaging over the multiple datasets. In
ImputeDB, we cannot assume the user desires multiple copies of the query result, though we
note that in further work, multiple imputation could be used within ImputeDB to reduce the
variance of estimates of some aggregates, like averages.

@@

It is worth noting that given the nature of the imputation strategy used here, imputation
operators become blocking in our system. An extension of ImputeDB could consider regression
algorithms that learn in an online nature; this would reduce the cost of the imputation at
the possible expense of quality. Again, given the imputation strategy-agnostic design of
ImputeDB, this would not pose a problem.

@@

\subsection{Complexity}
Our optimization algorithm builds off the approach taken by System R \cite{blasgen1981system}, therefore our algorithm still operates in exponential time. Indeed, 
note that if we remove our restriction on types of imputation($\delta_{min}, \mu_{min}, \mu_{max}$), and allow any arbitrary subset of attributes to be imputed,
then every single operator in the query plan has a number of imputations exponential in the number of dirty columns. Our restriction, instead, increases the number
of plans (in the worst case) at each operator by a factor of 3. Of course, this implies that in the worst case (where all dirty sets tracked are distinct throughout the query plan),
we explore a number of plans that further scales the number of plans in the original algorithm by an exponential factor.

### 2016-12-14 paper/experiments.tex
paper: Add some last edits

To evaluate the performance of our system, we generated plans for queries on two data sets: the American Community Survey and a synthetic dataset.
\subsection{Data sets} \label{subsec:datasets}

@@

We collected a set of queries (Figure~\ref{fig:queries}) that we think are both representative, in that they could reasonably be written by a user in the course of data analysis, and interesting to plan.

@@

\begin{table}
  \centerfloat
  \input{runtimes}
  \caption{Running time and error for queries with different imputation levels. Base error is the RMSE between the query run on clean data and the query run on dirty data without imputation.}
  \label{fig:experiments}
\end{table}

Figure~\ref{fig:experiments} provides a summary of our experimental results.
We run each query in four configurations: without imputation on clean data, without imputation on dirty data, and with imputation on dirty data for $\alpha = 0$ and $\alpha = 1$.
When a query returns a relation of the same ``shape'' (same number of tuples and same schema) on both clean and dirty data, we can compute an error value.
For our experiments, we report root-mean-square error (RMSE).
For imputations, we run queries using the two extreme values of $\alpha$: 0 and 1, representing a focus on runtime and on information loss, respectively.
For each query, the table displays the error between the query run on clean data and the query run on dirty data with imputation and the running time in seconds.

Planning times are not shown, as they are fractions of a second for all of the queries that we tested.
The overall low planning times highlight the practicality of our algorithm, despite the exponential complexity.
Running times clearly vary based on $\alpha$, allowing users to tune their query performance as desired.

@@

\begin{itemize}
\item For sum and count queries (\ref{q1}, \ref{q2}, \ref{q4}), imputation does a good job of reducing the error.
For each of these queries, we can reduce total error by several orders of magnitude.
Imputation performs very well in these cases, even for data which is missing uniformly at random, because each missing value directly increases the error of the result.

>>>>

To evaluate the performance of our system, we generate plans for queries on two data sets: the American Community Survey and a synthetic dataset.
\subsection{Data sets} \label{subsec:datasets}

@@

We collected a set of queries (Table~\ref{fig:queries}) that we think are both representative, in that they could reasonably be written by a user in the course of data analysis, and interesting to plan.

@@

\begin{table}
  \centerfloat
  \input{runtimes}
    \caption{Error ($\epsilon$, planning time ($t_p$), and running time ($t_r$)i or queries
    with different imputation levels. Base error is the RMSE between the query run on clean
    data and the query run on dirty data without imputation. \todo{Micah: Not clear to me
    what the symbol means.}}
  \label{fig:experiments}
\end{table}

Table~\ref{fig:experiments} provides a summary of our experimental results.
We run each query in four configurations: without imputation on clean data, without imputation on dirty data, and with imputation on dirty data for $\alpha = 0$ and $\alpha = 1$.
When a query returns a relation of the same ``shape'' (same number of tuples and same schema) on both clean and dirty data, we can compute an error value.
For our experiments, we report root-mean-square error (RMSE).
For imputations, we run queries using the two extreme values of $\alpha$: 0 and 1, representing a focus on runtime and on information loss, respectively.
For each query, the table displays the error between the query run on clean data and the query run on dirty data with imputation and the running time in seconds.

Planning times are not shown\todo{they are shown...}, as they are fractions of a second for all of the queries that we tested.
The overall low planning times highlight the practicality of our algorithm, despite the exponential complexity.
Running times clearly vary based on $\alpha$, allowing users to tune their query performance as desired.

@@

\todo{consider giving up on getting the references to work?}
\begin{itemize}
\item For sum and count queries (\ref{q1}, \ref{q2}, \ref{q4}), imputation does a good job of reducing the error.
For each of these queries, we can reduce total error by several orders of magnitude.
Imputation performs very well in these cases, even for data which is missing uniformly at random, because each missing value directly increases the error of the result.

### 2016-12-14 paper/introduction.tex
paper: Add some last edits

Handling incorrect or dirty data is a complex and challenging problem for data scientists.
One way in which a dataset can be dirty is for parts of it to be missing altogether.
Missing data is one of the simpler variants of dirty data, but if handled naively, can still cause analyses to be incorrect.
To handle this problem, users may manually clean their dataset by performing some regression analysis to replace missing data elements with likely values.
This process is called \emph{imputation}.
Traditional imputation methods replace all missing values in a dataset to create a new clean dataset.
Although manual imputation solves the problem of missing data, in the age of big data it may be very expensive to run an imputation algorithm on an entire dataset.
Additionally, it may not even be necessary to completely clean the data to make it usable.
Some users may be willing to run queries on dirty data, simply ignoring the missing values, as long as they do not have to pay the cost of imputation.
Others may want to run queries on a subset of the data, so do not need to impute every field in every record.

>>>>

Handling incorrect or dirty data is a complex and challenging problem for data scientists.
One way in which a dataset can be dirty is for parts of it to be missing altogether.
Missing data is one of the simpler variants of dirty data, but if handled naively, can still cause analyses to be incorrect.
To handle this problem, users may manually clean their dataset by performing
some statistical analysis to replace missing data elements with likely values.
This process is called \emph{imputation}.
Traditional imputation methods replace all missing values in a dataset to create a new clean dataset.
Although manual imputation solves the problem of missing data, in the age of big data it may be very expensive to run an imputation algorithm on an entire dataset.
Additionally, it may not even be necessary to completely clean the data to make it usable.
Some users may be willing to run queries on dirty data, simply ignoring the missing values, as long as they do not have to pay the cost of imputation.
Others may want to run queries on a subset of the data, so do not need to impute
every field in every record. Yet others may want to customize the 
imputation algorithm for the tradeoffs and demands of a particular domain.

### 2016-12-14 paper/relatedwork.tex
paper: Add some last edits

In \cite{akande2015empirical}, Akande et al analyze the performance of various
multiple imputation techniques on the American Community Survey dataset
(Sec~\ref{subsec:datasets}). The computational difficulties of imputing on large base
tables are well known and can limit approaches. For example, Akande finds that
one approaches (MI-GLM) crashes when attempting to impute on data that
includes variables with potentially large domains (ten categories specifically).
In contrast, \ProjectName{} allows users to specify a tradeoff between
information loss and computational complexity (in terms of time). Furthermore,
the query planner's imputation is guided by the requirements of each specific
query's operators, rather than requiring broad assumptions about query
workloads.

@@

While BayesDB can be used for value imputation, this step is not framed
within the context of query planning, but rather as an explicit step, using the
\verb|INFER| operation.

>>>>

In \cite{akande2015empirical}, Akande et al analyze the performance of various
multiple imputation techniques on the American Community Survey dataset
(Sec~\ref{subsec:datasets}). The computational difficulties of imputing on large base
tables are well known and can limit approaches. For example, Akande finds that
one approach (MI-GLM) is prohibitively expensive when attempting to impute on data that
includes variables with potentially large domains (ten categories in their case).
In contrast, \ProjectName{} allows users to specify a tradeoff between
information loss and computational complexity (in terms of time). Furthermore,
the query planner's imputation is guided by the requirements of each specific
query's operators, rather than requiring broad assumptions about query
workloads.

@@

While BayesDB can be used for value imputation, this step is not framed
within the context of query planning, but rather as an explicit statistical
inference step within the query language, using the \verb|INFER| operation.
